Bayesian calibration of a hybrid nuclear collision model\\using $\boldsymbol p$-Pb and Pb-Pb data from the LHC

### 2018-10-10 paper/substructure.tex
save revision work on the manuscript.

\title{
  Estimating initial state and quark-gluon plasma medium properties \\
  using a hybrid model with nucleon substructure \\
  calibrated to $\boldsymbol p$-Pb and Pb-Pb collisions at \texorpdfstring{$\mathbf{\sqrts=5.02}$}{}~TeV
}

@@

We model the QGP initial state in $p$-Pb and Pb-Pb collisions at $\sqrts=5.02$~TeV using a simple parametric form for boost-invariant entropy deposition employed by the \trento\ model \cite{Moreland:2014oya}.
Generally speaking, the initial three-dimensional distribution of matter produced in relativistic nuclear collisions is \emph{not} boost-invariant; longitudinal entropy deposition fluctuates both locally point-to-point in the transverse plane as well as globally event-by-event due to asymmetries in the sampled density of participant matter \cite{Ke:2016jrd, Bozek:2010vz}.
Nevertheless, boost-invariance has been shown to be a good approximation for both large and small collision systems when hydrodynamic observables are calculated from particles that are detected close to midrapidity \cite{Shen:2016zpp}.

The \trento\ model operates in the ultrarelativistic limit with a Lorentz factor $\gamma \gg 1$ such that each nucleus appears as a thin sheet of nuclear density in the laboratory frame.
The sheets of colliding nuclear density penetrate and pass through each other in proper time $\Delta \tau \approx D_\text{nucl} / (\gamma\, \beta_z)$ in the laboratory frame, where $D_\text{nucl}$ is the diameter of the nucleus in its rest frame, $\gamma$ is the usual Lorentz factor of the accelerated ions, and $\beta_z$ is their velocity along the beam axis.
The resulting nuclear overlap time $\Delta \tau \lesssim 0.1\ \fmc$ at top RHIC and LHC energies, and thus it is safe to neglect the initial transverse dynamics which occur while the nuclei pass through each other.
We therefore assume that the collision produces all secondary particles at uniform proper time $\tau = 0^+\ \fmc$, and that it deposits entropy (energy) at midrapidity which is a function of the locally varying beam integrated density of each nucleus.

Consider the collision of two protons $A$, $B$ with three-dimensional densities $\rho_{A,B}$ in their local rest frames.
The proton-proton overlap function
\begin{equation}
  \label{eq:tpp}
  T_{pp}(b) \equiv \int dx\, dy \int dz\, \rho_A(\x) \int dz\, \rho_B(\x + \mathbf{b}),
\end{equation}
describes the eikonal overlap of the two proton wave packets at fixed impact parameter $b$, where coordinates $(x, y)$ lie in the transverse plane, and $z$ is parallel to the beam axis.
Here we assume that each proton is comprised of smaller constituents---e.g.\ valence quarks, sea quarks, and small-x gluons---which may collide to produce secondary particles and contribute to the observed inelastic proton-proton cross section.

Within a picture of independent pairwise collisions between the constituents, a Glauber model model may be used to calculate the probability $P_\mathrm{coll}$ that the two protons collide inelastically at impact parameter $b$. In the limit when the number of constituents is large, it yields the particularly simple form
\begin{equation}
  \label{eq:pcoll}
  P_\mathrm{coll} = 1 - \exp[-\sigma_\mathrm{eff}\, T_{pp}(b)], \\
\end{equation}
where $\sigma_\mathrm{eff}$ is an effective cross section for pairwise inelastic collisions between the constituents, and $\sigma_{pp}^\mathrm{inel}$ is the total inelastic proton-proton cross section
\begin{equation}
  \label{eq:sigma_nn}
  \sigma_{pp}^\mathrm{inel} = \int 2 \pi b\, db\, P_\mathrm{coll}(b).
\end{equation}
The proton densities $\rho_{A,B}$ in Eq.~\eqref{eq:tpp} are commonly modeled using a spherically symmetric distribution.
For instance, the original implementation of the \trento\ model uses Gaussian protons, largely because it yields a simple analytic solution to Eq.~\eqref{eq:pcoll}.
Needless to say, such approximations are admittedly crude and may have a significant effect on the dynamics of small collision systems where the proton size is comparable to the size of the produced QGP medium.

A number of previous studies have investigated the effects of deformed or ``lumpy'' protons.
One common implementation is a superposition of three valence quarks, typically described by Gaussian or exponential form factors \cite{Welsh:2016siu, Bozek:2017jog, Schenke:2014zha, Schlichting:2014ipa, Adare:2015bua, Broniowski:2016pvx}.
The corresponding proton density $\rho(\mathbf{x})$ is then assumed to be that of predominantly small-x gluons, seeded by the distribution of color charge in each of the three valence quarks.

@@

In this work, we pursue a less restrictive and more parametric description of the proton where the number of substructure degrees of freedom are uncertain as depicted in Fig.~\ref{fig:substructure}.
We model each nucleon's density $\rho_{A,B}$ as a sum of $n_c$ independent constituents
\begin{equation}
  \label{eq:rho}
  \rho_{A, B}(\x) = \frac{1}{n_c} \sum\limits_{i=1}^{n_c} \rho_c(\mathbf{x} - \mathbf{x_i}),
\end{equation}
where each constituent density $\rho_c$ is described by a Gaussian distribution of width $v$
\begin{equation}
  \label{eq:constituent_density}
  \rho_c(\mathbf{x}) = \frac{1}{(2\pi v^2)^{3/2}} \exp\left(-\frac{\x^2}{2 v^2}\right),
\end{equation}
and each constituent's position $\mathbf{x_i}$ in Eq.~\eqref{eq:rho} is sampled from a Gaussian radial distribution with standard deviation $r$.
The effect of this additional nucleon substructure on the nuclear thickness functions is visible in Fig.~\ref{fig:thickness}

@@

Within the eikonal approximation, the initial entropy deposited at midrapidity and proper time $\tau=0^+\ \fmc$ is some function
\begin{equation}
  f: \T_A, \T_B \mapsto \frac{dS}{d^2x_\perp d\eta} \bigg\vert_{\eta=0},
\end{equation}
of the local density of participant matter $\T_A$, $\T_B$ in each nucleus.
A natural first guess for this mapping is the arithmetic mean
\begin{equation}
  \frac{dS}{d^2x_\perp d\eta}\,\bigg\vert_{\eta=0} \propto \frac{\T_A + \T_B}{2},
\end{equation}
which is simply a wounded nucleon model up to meaningless factor of two in the normalization.
The wounded nucleon model was in fact one of the first such mappings used as a proxy for initial particle production and entropy deposition in relativistic heavy-ion collisions \cite{Bialas:1976ed}.
It was quickly realized, however, that the wounded nucleon model predicts the wrong scaling for charged particle production as a function of collision centrality and hence the wrong scaling for initial entropy deposition as a function of participant thicknesses $\T_A$ and $\T_B$ \cite{Kharzeev:2000ph}.

>>>>

\title{
  Bayesian calibration of a hybrid nuclear collision model\\using $\boldsymbol p$-Pb and Pb-Pb data from the LHC
}

@@

We model the initial state of $p$-Pb and Pb-Pb collisions at $\sqrts=5.02$~TeV using the boost-invariant \trento\ model \cite{Moreland:2014oya}.
Generally speaking, the initial three-dimensional distribution of matter produced in relativistic nuclear collisions is \emph{not} boost-invariant; longitudinal entropy deposition fluctuates both locally point-to-point in the transverse plane as well as globally event-by-event due to asymmetries in the sampled density of participant matter \cite{Ke:2016jrd, Bozek:2010vz}.
Nevertheless, boost-invariance has been shown to be a good approximation for both large and small collision systems when hydrodynamic observables are calculated from particles that are detected close to midrapidity \cite{Shen:2016zpp}.

The \trento\ model operates in the ultrarelativistic limit with a Lorentz factor $\gamma \gg 1$ such that each nucleus appears as a thin sheet of nuclear density in the laboratory frame.
The sheets of colliding nuclear density penetrate and pass through each other in time $\Delta t \approx D_\text{nucl} / (\gamma\, \beta_z)$ in the laboratory frame, where $D_\text{nucl}$ is the diameter of the nucleus in its rest frame, $\gamma$ is the usual Lorentz factor of the accelerated ions, and $\beta_z$ is their velocity along the beam axis.
The resulting nuclear overlap time $\Delta t \lesssim 0.1\ \fmc$ at top RHIC and LHC energies, and thus it is safe to neglect the initial transverse dynamics which occur while the nuclei pass through each other.
We therefore assume that the collision produces all secondary particles at uniform proper time $\tau_0 \ll 1~\fmc$, and that it deposits energy at midrapidity which is a function of the locally varying beam integrated density of each nucleus.

Consider the collision of two protons, labeled $A$ and $B$, with three-dimensional densities $\rho_A$ and $\rho_B$ in their local rest frames.
The proton-proton overlap function
\begin{equation}
  \label{eq:tpp}
  T_{pp}(b) \equiv \int dx\, dy\, dz\, \rho_A(x, y, z)\, \rho_B(x - b, y, z),
\end{equation}
describes the eikonal overlap of the two proton wave packets at fixed impact parameter $b$, where coordinate $(x, y)$ lies in the transverse plane and $z$ is parallel to the beam axis.
Here we assume that each proton is comprised of smaller constituents---e.g.\ valence quarks, sea quarks, and small-x gluons---which may collide to produce secondary particles and contribute to the observed inelastic proton-proton cross section.

Within a picture of independent pairwise collisions between the constituents, a Glauber model model may be used to calculate the probability $P_\mathrm{coll}$ that the two protons collide inelastically at impact parameter $b$. In the limit when the number of constituents is large, it yields the particularly simple form
\begin{equation}
  \label{eq:pcoll}
  P_\mathrm{coll}(b) = 1 - \exp[-\sigma_\mathrm{eff}\, T_{pp}(b)], \\
\end{equation}
where $\sigma_\mathrm{eff}$ is an effective cross section for pairwise inelastic collisions between the constituents, and $\sigma_{pp}^\mathrm{inel}$ is the total inelastic proton-proton cross section
\begin{equation}
  \label{eq:sigma_nn}
  \sigma_{pp}^\mathrm{inel} = \int 2 \pi b\, db\, P_\mathrm{coll}(b).
\end{equation}
The proton densities $\rho_{A,B}$ in Eq.~\eqref{eq:tpp} are commonly modeled using a spherically symmetric distribution.
For instance, the original implementation of the \trento\ model uses Gaussian protons, largely because it yields a simple analytic solution to Eq.~\eqref{eq:pcoll}.
Needless to say, such approximations are admittedly crude and may have a significant effect on the dynamics of small collision systems where the proton size is comparable to the size of the produced QGP medium.

A number of previous studies have investigated the effects of deformed or ``lumpy'' protons.
One common implementation is a superposition of three valence quarks, typically described by Gaussian or exponential form factors \cite{Welsh:2016siu, Bozek:2017jog, Schenke:2014zha, Schlichting:2014ipa, Adare:2015bua, Broniowski:2016pvx}.
The corresponding proton density is then assumed to be that of predominantly small-x gluons, seeded by the distribution of color charge in each of the three valence quarks.

@@

In this work, we pursue a less restrictive and more parametric description of the proton where the number of substructure degrees of freedom are uncertain as depicted in Fig.~\ref{fig:substructure}.
We model each nucleon's three-dimensional density $\rho_{A,B}(\x)$ as a weighted sum of $n_c$ independent constituents,
\begin{equation}
  \label{eq:rho}
  \rho_{A, B}(\x) = \frac{1}{n_c} \sum\limits_{i=1}^{n_c} \frac{\gamma_i}{(2 \pi v^2)^{3/2}} \exp\left(-\frac{|\x - \x_i|^2}{2 v^2}\right),
\end{equation}
where each constituent density is assumed to be a Gaussian of width $v$, weighted by a gamma random variable $\gamma_i$ with unit mean and variance $1/k$.
These \emph{ad hoc} gamma random weights introduce additional multiplicity fluctuations in order to reproduce the large fluctuations observed in experimental proton-proton collisions.
We sample the constituent positions $\x_i$ independently (ignoring correlations) from a Gaussian radial distribution,
\begin{equation}
  \label{eq:radial_dist}
  P(\x_i) = \frac{1}{(2\pi r^2)^{3/2}} \exp\left(-\frac{|\x_i|^2}{2 r^2}\right),
\end{equation}
parametrized by a sampling radius $r$ which varies the average separation of the constituents.
Figure~\ref{fig:thickness} shows the effect of this additional nucleon substructure on the nuclear thickness functions.

@@

The initial energy density $e$ deposited at midrapidity and proper time $\tau_0 \approx 0~\fmc$ is some function
\begin{equation}
  e(\x_\perp, \eta_s=0, \tau_0) = f(\T_A, \T_B)
\end{equation}
of the local density of participant matter $\T_A$, $\T_B$ in each nucleus.
A natural first guess for this mapping is the arithmetic mean
\begin{equation}
  f = \text{const} \times \frac{\T_A + \T_B}{2},
\end{equation}
which is simply a wounded nucleon model up to meaningless factor of two in the normalization.
The wounded nucleon model was in fact one of the first such mappings used as a proxy for initial particle production and entropy deposition in relativistic heavy-ion collisions \cite{Bialas:1976ed}.
It was quickly realized, however, that the wounded nucleon model predicts the wrong scaling for charged particle production as a function of collision centrality and hence the wrong scaling for initial entropy deposition as a function of participant thicknesses $\T_A$ and $\T_B$ \cite{Kharzeev:2000ph}.
